Unlocking the Potential of microRNAs: Machine Learning Identifies Key Biomarkers for Myocardial Infarction Diagnosis

### 2023-08-07 Complete-V1/Complete-V1.Rmd
The version after revision

---
title: Machine Learning Deriven Set of microRNAs as a Novel Biomarker for Myocardial Infarction Diagnosis
#titlerunning: Short form of title (if too long for head)
authorrunning: M. Samadishadlou et al.
#thanks: | 
#    Grants or other notes about the article that should go on the front 
#    page should be placed here. General acknowledgments should be placed at the
#    end of the article.

authors: 
- name: Mehrdad Samadishadlou
  address: Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran
  email: samadishadlou@tbzmed.ac.ir

- name: Reza Rahbarghazi
  address: Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran
  email: rahbarghazir@tbzmed.ac.ir

@@

keywords:
- microRNA
- Machine Learning
- Myocardial Infarction
- Bioinformatics

@@

abstract: |
  MicroRNAs (miRNAs) play a crucial role in regulating adaptive and maladaptive responses in cardiovascular diseases, making them attractive targets for potential therapeutics. However, their potential as novel biomarkers for diagnosing cardiovascular diseases requires systematic evaluation. In this study, we aimed to identify a key set of miRNA biomarkers using integrated bioinformatics and machine learning analysis. We combined and analyzed three gene expression datasets from the Gene Expression Omnibus (GEO) database, which contains peripheral blood mononuclear cell samples from individuals with myocardial infarction (MI), stable coronary artery disease (CAD), and healthy individuals. Additionally, we selected a set of miRNAs based on their area under the curve of the receiver operating characteristic curve (AUC-ROC) to separate the CAD/MI samples. We designed a two-layer architecture for sample classification, with the first layer isolating healthy samples from not-healthy ones, and the second layer classifying stable CAD and MI samples. We trained different machine learning models using both biomarker sets and evaluated their performance on a test set. We identified miR-21, miR-186, and miR-32 as the only miRNAs among the differentially expressed genes, and a set including miR-186, miR-21, miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by their AUC-ROC. Both biomarker sets could distinguish healthy from not-healthy samples with complete accuracy. The best performance for the classification of CAD and MI was achieved with an SVM model trained using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.95, and an accuracy of 0.88. Our study demonstrated that miRNA signatures derived from peripheral blood could serve as valuable novel biomarkers for cardiovascular diseases.

@@

bibliography: [ForArticleRefrence.bib, discussion.bib, Introduction.bib, mir296.bib]
biblio-style: spmpsci
# bibstyle options spbasic(default), spphys, spmpsci
---

```{r setup, include=FALSE, cache.rebuild=TRUE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, captioner, bundesligR, stringr, kableExtra)
```

@@

At present, cardiovascular diseases (CVDs) are the leading cause of human mortality with 32% of all global deaths. It is estimated that about 85% of CVDs mortality were diagnosed with myocardial infarction (MI) [@CVD]. MI is an acute coronary syndrome with sudden blockage and stenosis of the coronary artery, and subsequent myocardial ischemia, leading to extensive cardiomyocyte damage and necrosis [@Macro].

Over the last 50 years, numerous attempts have been collected to use biomarkers to facilitate diagnosis, assess risk, follow-up therapy, and determine therapeutic efficacy in CVDs candidates. Based on the released guidelines, cardiac troponins (cTns) are used as a highly-sensitive and accurate approach for the detection of MI. Despite the inherent advantages, the high-rate sensitivity of cTn-based assays has also led to more false positive results [@17], which do necessitate the advent and development of new modalities with pathological values. To improve diagnostic value upon existing MI biomarkers, the combination of complementary biological markers, such as microRNAs (miRNAs) and other genetic factors, is proposed. Previous researches support the notion that miRNAs exhibit the great potential to be used as alternative biomarkers in CVDs detection and follow-up [@miR]. It is suggested that miRNAs possess 18-22 nucleotides and can play a crucial role in the regulation of gene expression. Evidence point to the fact that miRNAs are involved in the pathogenesis of cardiac tissue injury and could be considered as theranostics targets in terms of CVDs. These elements can easily circulate in biofluids with both diagnostic as well as prognostic values [@m1]. Several biological activities such as angiogenesis, cardiomyocyte growth and contractility, lipid metabolism, plaque formation, and cardiac rhythm are regulated by miRNAs [@m2]. It is postulated that the function and diagnostic properties of miRNAs are beyond the myocardium in CVD patients. To be specific, the expression of miRNAs can vary in different biofluids and cell components such as serum and peripheral blood mononuclear cells (PBMCs) [@36].

PBMCs are a fraction of white blood cells (WBCs), including monocytes, lymphocytes, macrophages, and other cells belonging to the immune system [@PBMC-miR]. Emerging data have indicated that PBMCs can be used as a valid source of biomarkers for monitoring various pathological conditions. Of note, the alteration of mRNAs and miRNAs under pathological conditions gives us valuable information about different kinds of disorders. PBMCs could recapitulate the conditions of the target tissues, thus, providing a highly sensitive and specific source of biomarkers [@Meder6]. Commensurate with these conditions, these cells are repositories of dysregulated genes and miRNAs expression profiles in CVDs [@PBMC-miR; @Meder6].

In recent years, the advent and use of machine learning (ML) is an exciting prospect for advancing scientific discoveries. Although the concept of ML and its initial algorithms were conceived many years ago, recent improvements in computing power and access to vast amounts of data have shown that ML techniques outperform classical statistical methods in various fields. Furthermore, the progress made in omics technologies has enabled the analysis of massive and intricate biological data sets, consisting of hundreds to thousands of samples, which makes it possible for ML to extract valuable biological information from such data [@ML-Bio]. Therefore, ML offers novel techniques to integrate and analyze the various omics data enabling the discovery of de novo biomarkers. These biomarkers help us in accurate disease prediction, patient stratification, and finding new therapeutics [@ML].

In this study, we aimed to identify potential miRNA biomarkers for MI patients by combining and analyzing three different microarray datasets from PBMCs. It is suggested that the integration of omics data with bioinformatics and ML techniques could be a promising tool in the discovery of new and more accurate biomarkers for monitoring MI. Besides, this approach can deepen our vision into the underlying mechanisms of MI and aid in the development of valid diagnostic biomarkers, and patient stratification.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO) database (<https://www.ncbi.nlm.nih.gov/geo/>). To obtain sufficient classification power between MI, healthy and stable coronary artery disease (CAD) samples, a relatively large sample size was required. Therefore, GSE59867 for MI and CAD samples, and GSE56609 and GSE54475 for healthy samples were selected. All samples were produced using Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. Only healthy, stable CAD and early-stage MI samples were selected from these datasets for further analyses. The basic information for the three datasets evaluated in the current study is provided in Table \@ref(tab:datasets).

@@

kbl(df, format = "markdown", booktabs = T, align = "c", linesep = c(""), caption = "Basic information of the GEO microarray datasets.")
```

## Pre-processing

Raw data (CEL files) of all datasets were downloaded from the GEO and pre-processed using the fRMA package [@fRMA]. fRMA allowed to pre-process of individual microarray samples and combining them consistently for analysis. For each dataset, background correction was performed using the RMA algorithm and then it was quantile normalized based on the reference distribution. During summarization, batch effects were removed and variances of the gene expressions were estimated by taking into account these probe-specific effects. For those multiple probe sets matched to the identical gene, the mean log fold change was retained. Therefore, fRMA can be seen as a batch effect removal technique for different datasets that are produced by identical microarray platforms. Thus, to ensure batch effect removal, the principal component analysis and the relative log expression of train samples were plotted before and after fRMA [@BER].

@@

The barcode algorithm proposed by McCall et al. [@Barcode] transformed the actual expression values into binary barcode values. Huge sets of samples were collected and normalized using fRMA for several platforms as well as for Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. The distribution of the expressed and unexpressed observed intensities for each gene is estimated using these normalized sets. Genes were considered expressed (and their value coded to 1) or unexpressed (and their value coded to 0) according to the following equation:

@@

where $x_{ij}$ is the normalized intensity of gene $i$ in sample $j$, $C$ is a user-defined parameter, $\sigma^{ne}$ is the standard deviation of the non-expressed distribution, and $\mu^{ne}$ is the mean of the non-expressed distribution. The barcode representation of a sample is a vector of ones and zeros denoting which genes are estimated to be expressed (ones) and unexpressed (zeros). The barcode algorithm was implemented by the barcode function in the R fRMA package, and the default value of $C$ was used.

To determine if the expressed ratios differed in the MI group versus the healthy control group, Fisher's exact test for individual genes was carried out upon the barcode values. Genes with a false discovery rate (FDR) of $< 0.05$, which was calculated through the Benjamini-Hochberg (BH) procedure to adjust for multiple testing issues, were considered as differentially expressed genes. The same procedures were conducted on CAD versus healthy controls as well as MI versus CAD group to find the DEGs between them.

@@

Using the R clusterProfiler package, the Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analysis and Gene Ontology (GO) functional annotation were carried out on the differentially expressed genes. The GO analysis included biological process (BP), cellular component (CC) and molecular function (MF) categories. An adjusted p-value of less than 0.05 was considered to indicate a statistically significant difference. Enrichments were conducted on the MI-healthy and CAD-healthy DEGs. In these analyses, all default parameters were used.

@@

The ML analysis was performed using Python software, ver. 3.9, Numpy [@numpy], pandas [@pandas], and Scikit-Learn packages [@scikit-learn]. Whenever hyper-tuning was needed, the scikit-opt package [@scikitopt] was used. In all ML analyses, the datasets were divided into train and test sets by a 0.7:0.3 ratio and all reported results are the average of 10-fold cross-validation.

Two different approaches were used for selecting miRNAs for model training. The first approach was using the miRNAs that are differentially expressed. In the second approach, miRNAs with individual area under the curve of the receiver operating characteristic (AUC-ROC) over 0.8 for separating MI from CAD were selected. Having the result of these two different approaches can provide an informative comparison between the predictive capabilities of sets of miRNAs selected with different logics.

@@

In this approach, a two-layer architecture was deployed to the data to maximize the prediction values. The first layer predicted whether a sample is healthy or not, and the second layer separated MI from CAD in the samples which were predicted as not healthy in the first layer. To this end, a distinct ML model was trained for each layer. Since there is a limited number of miRNAs in DEGs, both layers were trained with all of them. For further comparison with the models' performance, the ROC curve of each miRNA for classifying healthy and not-healthy, as well as CAD and MI, were generated using a Logistic Regression model.

#### First layer for isolation of healthy and not-healthy samples:

An SVM model using RBF kernels was trained and hyper-tuned using all miRNAs in DEGs. To handle the severe imbalance in the number of samples (51 for the healthy and 157 for the not-healthy groups), the sample weight for the healthy and the not-healthy samples were set to 1 and 0.5, respectively.

#### Second layer for seperating MI and CAD samples:

For the sake of reaching the highest classification performance using the set of miRNAs, different models were investigated. To do so, SVM (with linear, polynomial, and RBF kernels), Logistic Regression (LR), Random Forests (RF), k-Nearest Neighbor (kNN), Gradient Boosting (GB), XGBoost (XGB) and Decision Tree (DT) models were trained. All models were trained with their pre-set parameters with 10-fold cross-validation. The criteria for choosing the best model were the highest accuracy and AUC-ROC on the test set. The best model was hyper-tuned with the scikit-opt package [@scikitopt] to get the best predictive performance.

@@

Like the previous approach, a two-layers strategy was conducted. The first layer classified samples into healthy and not-healthy, and the separated MI and CAD samples. However, to keep the number of miRNAs as low as possible miRNAs were selected from the second layer, and then their performance wase evaluated in the first layer. AUC-ROC of all miRNAs for classifying MI and CAD samples were calculated and the miRNAs with the AUC-ROC over 0.8 were selected. The ROC curves for each selected miRNA for separating healthy samples from not-healthy and MI from CAD samples were also plotted for further comparison.

#### First layer for the detection of healthy and not-healthy samples:

An SVM model with an RBF kernel was trained using the selected set of miRNAs. Additionally, the model was hyper-tuned to find the hyper-parameters for the highest AUC-ROC and accuracy. The ROC curve and confusion matrix for the best model were reported.

#### Second layer for separating MI and CAD:

The selected miRNAs set was used to train different algorithms to find the best model. Similar to the previous approach, SVM (with linear, polynomial, and RBF kernels), LR, RF, kNN, GB, XGB, and DT were trained. All models were trained with their pre-set parameters using 10-fold cross-validation. The models with the highest AUC-ROC and accuracy on the test set were selected and hyper-tuned using the scikit-opt package [@scikitopt]. The ROC curve and confusion matrix for the best model were reported.

@@

## Pre-processing

The PCA plots of the training samples are shown in Figure \@ref(fig:PCA)A and B. As shown, healthy samples were separated from CAD or MI samples in primary data and also after conducting fRMA. In the RLE plot, there was a distinct difference between dataset means for all samples before conducting fRMA (Figure \@ref(fig:PCA)C). All datasets were rearranged around 0 in the RLE plot after conducting fRMA (Figure \@ref(fig:PCA)D). Moreover, there was a clear change in inter-quantile distances, but the values still were over 0.1.

```{r PCA, echo=F, out.width="100%", fig.cap="PCA and RLE plot for all samples before and after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

@@

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between the MI and the healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in MI compared to the healthy controls. In CAD and healthy groups comparison, we found `r dim(C_H)[1]` DEGs, of which `r sum(C_H$logFC > 0)` and and `r sum(C_H$logFC < 0)` DEGs were up- and down-regulated, respectively in CAD samples. In the MI and CAD groups, the number of DEGs was `r dim(M_C)[1]`, and the number of up- and down-regulated genes in MI samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)`, respectively in comparison with CAD samples. These data are summarized in Table \@ref(tab:DEGstab).

```{r venn, echo=F, out.width="45%", fig.align='center', fig.cap="Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD comparison."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/venny.png")
```

The Venn diagram in Figure \@ref(fig:venn) shows that CAD and MI samples share a majority of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs of CAD/healthy, 531 genes were common which is 62% of MI/healthy DEGs and 79% of CAD/healthy DEGs.

## Gene ontology (GO) and Kyoto Encyclopedia of Genes and Genomes (KEGG) enrichment analyses of the DEGs.

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on MI-healthy and CAD-healthy DEGs. In MI versus healthy samples, GO enrichment analysis in the BP category, suggested that the DEGs were enriched in "immune response-regulating signaling pathway", "lymphocyte differentiation", "immune response-regulating cell surface receptor signaling pathway", and "leukocyte activation involved in immune response" (Figure \@ref(fig:MIHEnrich)A). In the CC category, the DEGs were enriched in "secretory granule membrane", "azurophil granule", "ficolin-1-rich granule", "tertiary granule", and "ficolin-1-rich granule membrane" (Figure \@ref(fig:MIHEnrich)B). In the MF category, the DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Figure \@ref(fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway", "Lipid and atherosclerosis", and "Hematopoietic cell lineage" (Figure \@ref(fig:MIHEnrich)D).

@@

kbl(df, booktabs = T, align = "c",linesep = c(""), digits = c(1, 60), caption = "Target miRNAs log fold-change and adjusted p-values for CAD samples relative to healthy, MI samples relative to healthy, and MI samples relative to CAD.") %>%
  kable_styling(full_width = T) %>%
  add_header_above(c(" " = 1, "CAD/Healthy" = 2, "MI/Healthy" = 2, "MI/CAD" = 2))
```

```{r Expall, echo=F, out.width="90%", fig.align='center', fig.cap="Expression profile of all miRNAs in two approach in different sample classes."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/Expressionforall.pdf")
```

>>>>

---
title: "Unlocking the Potential of microRNAs: Machine Learning Identifies Key Biomarkers for Myocardial Infarction Diagnosis"
#titlerunning: Short form of title (if too long for head)
authorrunning: M. Samadishadlou et al.
#thanks: | 
#    Grants or other notes about the article that should go on the front 
#    page should be placed here. General acknowledgments should be placed at the
#    end of the article.

authors: 
- name: Mehrdad Samadishadlou
  address: Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran

- name: Reza Rahbarghazi
  address: Medical Nanotechnology gruop, Tabriz University of Medical Sciences, Tabriz, Iran

- name: Zeynab Piryaei 
  address: Laboratory of Complex Biological Systems and Bioinformatics (CBB), Department of Bioinformatics, Institute of Biochemistry and Biophysics (IBB), University of Tehran, Iran.

- name: Mahdad Esmaeili
  address: Medical Bioengineering Department, School of Advanced Medical Sciences, Tabriz University of Medical Sciences, Tabriz, Iran

- name: Çığır Biray Avcı
  address: Medical Biology Department, School of Medicine, Ege University, İzmir, Turkey

@@

- name: Kaveh Kavousi
  address: Laboratory of Complex Biological Systems and Bioinformatics (CBB), Department of Bioinformatics, Institute of Biochemistry and Biophysics (IBB), University of Tehran, Iran
  email: kkavousi@ut.ac.ir

keywords:
- microRNA
- Machine Learning
- Myocardial Infarction
- Bioinformatics
- Biomarker

@@

abstract: |
  MicroRNAs (miRNAs) play a crucial role in regulating adaptive and maladaptive responses in cardiovascular diseases, making them attractive targets for potential biomarkers. However, their potential as novel biomarkers for diagnosing cardiovascular diseases requires systematic evaluation. In this study, we aimed to identify a key set of miRNA biomarkers using integrated bioinformatics and machine learning analysis. We combined and analyzed three gene expression datasets from the Gene Expression Omnibus (GEO) database, which contains peripheral blood mononuclear cell (PBMC) samples from individuals with myocardial infarction (MI), stable coronary artery disease (CAD), and healthy individuals. Additionally, we selected a set of miRNAs based on their area under the receiver operating characteristic curve (AUC-ROC) for separating the CAD and MI samples. We designed a two-layer architecture for sample classification, in which the first layer isolates healthy samples from unhealthy samples, and the second layer classifies stable CAD and MI samples. We trained different machine learning models using both biomarker sets and evaluated their performance on a test set. We identified miR-21, miR-186, and miR-32 as the only miRNAs among the differentially expressed genes, and a set including miR-186, miR-21, miR-197, miR-29A, and miR-296 as the optimum set of miRNAs selected by their AUC-ROC. Both biomarker sets could distinguish healthy from not-healthy samples with complete accuracy. The best performance for the classification of CAD and MI was achieved with an SVM model trained using the biomarker set selected by AUC-ROC, with an AUC-ROC of 0.96 and an accuracy of 0.94 on the test data. Our study demonstrated that miRNA signatures derived from PBMCs could serve as valuable novel biomarkers for cardiovascular diseases.

@@

bibliography: [For Article Refrence.bib, discussion.bib, Introduction.bib, mir296.bib]
biblio-style: spbasic
# bibstyle options spbasic(default), spphys, spmpsci
---

```{r setup, include=FALSE, cache.rebuild=TRUE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(knitr, captioner, bundesligR, stringr, kableExtra, ggplot2, ggvenn)
```

@@

Cardiovascular diseases (CVDs) are the leading cause of human mortality, accounting for 32% of all global deaths. It is estimated that approximately 85% of CVD mortality is due to myocardial infarction (MI) [@CVD]. MI is an acute coronary syndrome characterized by sudden blockage and stenosis of the coronary artery and subsequent myocardial ischemia, leading to extensive cardiomyocyte damage and necrosis [@Macro].

Over the last 50 years, numerous attempts have been made to use biomarkers to facilitate diagnosis, assess the risk, follow-up therapy, and determine therapeutic efficacy in CVD candidates. Based on released guidelines, cardiac troponins (cTns) are used as a highly sensitive and accurate approach for detecting MI. Despite these inherent advantages, the high sensitivity of cTn-based assays has also led to more false-positive results [@17], necessitating the advent and development of new modalities with pathological value. To improve the diagnostic value of  existing MI biomarkers, a combination of complementary biological markers, such as microRNAs (miRNAs) and other genetic factors, has been proposed. Previous research supports the notion that miRNAs exhibit great potential as alternative biomarkers for CVD detection and follow-up [@miR]. It has been suggested that miRNAs possess 18-22 nucleotides and play a crucial role in the regulation of gene expression. Evidence indicates that miRNAs are involved in the pathogenesis of cardiac tissue injury [@m1]. Several biological processes, such as angiogenesis, cardiomyocyte growth and contractility, lipid metabolism, plaque formation, and cardiac rhythm, are regulated by miRNAs [@m2]. These elements can easily circulate in biofluids and could be considered theranostic targets in terms of CVDs [@m1].  It has been postulated that the function and diagnostic properties of miRNAs are beyond the myocardium in patients with CVD.  Specifically, the expression of miRNAs can vary in different biofluids and cell components such as serum and peripheral blood mononuclear cells (PBMCs) [@36].

PBMCs are a fraction of white blood cells, including monocytes, lymphocytes, macrophages, and other cells of the immune system [@PBMC-miR]. Emerging data indicate that PBMCs can be used as a valid source of biomarkers for monitoring various pathological conditions. Of note, the alteration of mRNAs and miRNAs under pathological conditions provides valuable information about different kinds of disorders. PBMCs can recapitulate the conditions of target tissues, thus providing a highly sensitive and specific source of biomarkers [@Meder6]. Combined with these conditions, these cells are repositories of dysregulated genes and miRNA expression profiles in CVDs [@PBMC-miR; @Meder6].

In recent years, the advent and application of machine learning (ML) has been an exciting prospect for advancing scientific research. Although the concept of ML and its initial algorithms were conceived many years ago, recent improvements in computing power and access to vast amounts of data have demonstrated that ML techniques outperform classical statistical methods in various fields. Furthermore, the progress made in omics technologies has enabled the analysis of massive and intricate biological datasets, consisting of hundreds to thousands of samples, which makes it possible for ML to extract valuable biological insights and information from such data [@ML-Bio]. Consequently, ML provides innovative methods for merging and interpreting diverse types of omics data, leading to the identification of new biomarkers. These biomarkers can aid in precise disease prediction, patient stratification, and the development of novel therapeutic approaches [@ML].

In this study, we aimed to identify potential miRNA biomarkers in patients with MI by combining and analyzing three different microarray datasets from PBMCs. The integration of omics data with bioinformatics and ML techniques could be a promising tool in the discovery of new and more accurate biomarkers for monitoring MI. Additionally, this approach can deepen our understanding of the underlying mechanisms of MI and aid in the development of valid diagnostic biomarkers and patient stratification.

@@

Microarray datasets were obtained from the Gene Expression Omnibus (GEO) database (<https://www.ncbi.nlm.nih.gov/geo/>). To obtain sufficient classification power between MI, healthy, and CAD samples, a relatively large sample size is needed Therefore, GSE59867 for the MI and CAD samples and GSE56609 and GSE54475 for the healthy samples were selected. All samples were produced using Affymetrix Human Gene 1.0 ST Array platform (GPL6244). Only healthy, CAD, and early-stage MI samples were selected from these datasets for further analysis. The basic information for the three datasets evaluated in this study is provided in Table \@ref(tab:datasets). Bioinformatics analyses were conducted using R, ver. 4.2.0 [@R], RStudio [@RStudio]. All plots and graphics of these sections were created using the ggplot2 R package [@ggplot].

@@

kbl(df, format = "markdown", booktabs = T, align = "c", linesep = c(""), caption = "Sample information on the GEO microarray datasets.")
```

## Preprocessing

The raw data in the form of CEL files from all datasets were obtained from GEO. To prepare the data for analysis, we utilized the fRMA package [@fRMA] to facilitate preprocessing of individual microarray samples and their consistent combination. For each dataset, background correction was applied using the RMA algorithm, followed by quantile normalization based on the reference distribution. To account for probe-specific effects, batch effects were eliminated during summarization and gene expression variances were estimated accordingly. In cases where multiple probe sets matched the same gene, the mean log-fold change was retained. Consequently, fRMA can serve as a technique to remove batch effects across diverse datasets generated by identical microarray platforms [@BER]. To ensure the effectiveness of the batch effect removal, we employed principal component analysis (PCA) and relative log expression (RLE) plots to visualize the data before and after applying fRMA.

@@

The barcode algorithm was introduced by McCall et al. [@Barcode], aimed to convert actual expression values into binary barcode values. Extensive sample collections were gathered and normalization was performed using fRMA across multiple platforms, including the Affymetrix Human Gene 1.0 ST Array (GPL6244) platform. By utilizing these normalized datasets, the distribution of the observed intensities for both the expressed and unexpressed genes was estimated. The determination of whether a gene was expressed or not was based on the following equation, where a value of 1 indicates expression and a value of 0 indicates non-expression:

@@

In the barcode algorithm, the normalized intensity of gene $i$ in sample $j$ is denoted as $x_{ij}$. A user-defined parameter, $C$, was introduced along with the standard deviation ($\sigma^{ne}$) and mean ($\mu^{ne}$) of the non-expressed distribution. Based on these values, the barcode representation of a sample was generated as a vector consisting of ones and zeros, representing the estimated expression (ones) and non-expression (zeros) of each gene. The barcode function within the R fRMA package was employed to implement the barcode algorithm, utilizing the default value of $C$.

To assess the differences in expressed ratios between the MI and healthy control groups, Fisher's exact test was performed on the barcode values of individual genes. Genes that exhibited a false discovery rate (FDR) below $0.05$, calculated using the Benjamini-Hochberg procedure to account for multiple testing issues were identified as differentially expressed genes (DEGs). The same procedures were applied to the CAD versus healthy control comparison, as well as to the MI versus CAD group, to identify DEGs specific to each comparison.

@@

The R clusterProfiler package [@cluster] was utilized to perform Kyoto Encyclopedia of Genes and Genomes (KEGG) pathway enrichment analysis and Gene Ontology (GO) functional annotation on the set of DEGs. GO analysis encompassed three categories: biological process (BP), cellular component (CC), and molecular function (MF). For statistical significance, an adjusted p-value threshold of less than 0.05 was employed. Enrichment analyses were conducted separately for DEGs specific to the MI-healthy and CAD-healthy comparisons. All the default parameters provided by the package were used in the analyses.

@@

ML analysis was performed using Python software, ver. 3.9, Numpy [@numpy], Pandas [@pandas], and Scikit-Learn packages [@scikit-learn]. Whenever hypertuning was needed, the Scikit-opt package [@scikitopt] was used. In all ML analyses, the datasets were divided into training and test sets at a 0.7:0.3 ratio, and all reported results are the average of 10-fold cross-validation.

Two different approaches were used to select miRNAs for model training. The first approach was to use differentially expressed miRNAs. In the second approach, miRNAs with individual AUC-ROCs over 0.8 for separating MI from CAD were selected. The results of these two approaches can provide an informative comparison between the predictive capabilities of sets of miRNAs selected with different logics.

@@

In this approach, a two-layer architecture is deployed to the data to maximize the prediction values. The first layer predicted whether a sample was healthy or not, and the second layer separated MI from CAD in the samples that were predicted as not healthy in the first layer. To this end, a distinct ML model was trained for each layer. Because, there were a limited number of miRNAs in the DEGs, both layers were trained with all of them. For further comparison with the models' performance, the ROC curve of each miRNA for classifying healthy and not-healthy, as well as CAD and MI, was generated using a logistic regression model.

#### First layer for the isolation of healthy and not-healthy samples:

A support vector machine (SVM) model using RBF kernels was trained and hypertuned using all miRNAs in the DEGs. To handle the severe imbalance in the number of samples (51 for the healthy group and 157 for the not-healthy group), the sample weights for the healthy and the not-healthy samples were set to 1 and 0.5, respectively. The ROC curve and confusion matrix for the model are reported.

#### Second layer for separating the MI and CAD samples:

Different models were investigated to achieve the highest classification performance. To do so, SVM (with linear, polynomial, and RBF kernels), logistic regression (LR), random forests (RF), k-nearest neighbor (kNN), gradient boosting (GB), XGBoost (XGB) and decision tree (DT) models were trained. All models were trained with their preset parameters using 10-fold cross-validation. The criteria for selecting the best model were the highest accuracy and AUC-ROC for the test set. The best model was hypertuned using the scikit-opt package [@scikitopt] for the best classification performance. The ROC curve and confusion matrix for the best model are reported.

@@

As in the previous approach, a two-layer strategy was employed. The first layer classified samples into healthy and not-healthy, and the second layer separated the MI and CAD samples. However, to keep the number of miRNAs as low as possible, miRNAs were selected from the second layer and their performance was evaluated in the first layer. The AUC-ROC of all miRNAs for classifying MI and CAD samples was calculated, and miRNAs with AUC-ROC $> 0.8$ were selected. ROC curves for each selected miRNA for separating healthy samples from not-healthy samples and MI from CAD samples were also plotted for further comparison.

#### First layer for the isolation of healthy and not-healthy samples:

An SVM model with an RBF kernel is trained using the selected set of miRNAs. Additionally, the model was hypertuned to find the hyperparameters for the highest AUC-ROC and accuracy. The same sample weights as in the previous approach (1 for healthy and 0.5 for not-healthy samples) were used. The ROC curve and confusion matrix for the model were reported.

#### Second layer for separating the MI and CAD samples:

The selected miRNA set was used to train different algorithms to determine the best model. Similar to the previous approach, the SVM (with linear, polynomial, and RBF kernels), LR, RF, kNN, GB, XGB, and DT models were trained. All models were trained with their preset parameters using 10-fold cross-validation. The models with the highest AUC-ROC and accuracy on the test set were selected and hypertuned using the scikit-opt package [@scikitopt]. The ROC curve and confusion matrix for the best model were reported.

@@

## Preprocessing

The PCA plots of the samples are shown in Figures \@ref(fig:PCA)A and B. Healthy samples were separated from the CAD or MI samples in the primary data and after conducting fRMA. In the RLE plot, there was a distinct difference between the dataset means for all samples before fRMA was performed (Figure \@ref(fig:PCA)C). All datasets were rearranged to approximately 0 in the RLE plot after fRMA was conducted (Figure \@ref(fig:PCA)D). Moreover, there was an apparent change in the interquantile distances, but the values were still greater than 0.1.

```{r PCA, echo=F, out.width="100%", fig.cap="Principal component analysis plots for (A) primary data and (B) the data after fRMA, and the relative log expression plots for (C) primary data and (D) the data after fRMA."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Pics/PCA and RLE of trains.pdf")
```

@@

According to the cutoff criterion of $FDR < 0.05$, there were `r dim(M_H)[1]` DEGs between MI and healthy samples. Among them, `r sum(M_H$logFC > 0)` were up-regulated, and `r sum(M_H$logFC < 0)` were down-regulated in the MI group compared to the healthy group. In the CAD and healthy group comparison, we found `r dim(C_H)[1]` DEGs, of which `r sum(C_H$logFC > 0)` and `r sum(C_H$logFC < 0)` DEGs were up- and down-regulated, respectively, in CAD samples. In the MI and CAD groups, the number of DEGs was `r dim(M_C)[1]`, and the numbers of up- and down-regulated genes in MI samples were `r sum(M_C$logFC > 0)` and `r sum(M_C$logFC < 0)`, respectively, compared to CAD samples. The data are summarized in Table \@ref(tab:DEGstab).

```{r venn, echo=F, out.width="45%", fig.align='center', fig.cap="Venn diagram for DEGs in CAD/Healthy, MI/Healthy, and MI/CAD."}

ch <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs CAD-Healthy.txt", check.names = F)$DEGs

mh <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs MI-Healthy.txt", check.names = F)$DEGs

mc <- read.table("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/DEG and DEM/Fisher results/DEGs MI-CAD.txt", check.names = F)$DEGs

df <- list(`CAD/Healthy` = ch, `MI/Healthy` = mh, `MI/CAD` = mc)

ggvenn(df, show_percentage = F, stroke_alpha = 0.5, fill_alpha = 0.4, text_size = 6, set_name_size = 6) + theme(text = element_text(family = "Times"))

```

The Venn diagram in Figure \@ref(fig:venn) shows that the CAD and MI samples shared most of their DEGs. From 860 DEGs of MI/healthy and 670 DEGs of CAD/healthy, 531 genes were common, which is 62% of MI/healthy DEGs and 79% of CAD/healthy DEGs.

## GO and KEGG enrichment analyses of the DEGs

To explore the biological classification of the DEGs, we performed GO and KEGG pathway enrichment analyses on the MI/healthy and CAD/healthy DEGs. For MI/healthy, GO enrichment analysis in the BP category suggested that the DEGs were enriched in "immune response-regulating signaling pathway," "lymphocyte differentiation," "immune response-regulating cell surface receptor signaling pathway," and "leukocyte activation involved in immune response" (Figure \@ref(fig:MIHEnrich)A). In the CC category, DEGs were enriched in "secretory granule membrane," "azurophil granule," "ficolin-1-rich granule," "tertiary granule," and "ficolin-1-rich granule membrane" (Figure \@ref(fig:MIHEnrich)B). In the MF category, DEGs were involved in "cadherin binding" and "MHC class I protein binding" (Figure \@ref(fig:MIHEnrich)C). KEGG pathway analysis indicated that the DEGs were related to the following pathways: "Chemokine signaling pathway," "Lipid and atherosclerosis," and "Hematopoietic cell lineage" (Figure \@ref(fig:MIHEnrich)D).

@@

kbl(df, booktabs = T, align = "c",linesep = c(""), digits = c(1, 60), caption = "Investigated miRNAs log fold-change and adjusted p-values for CAD samples relative to healthy, MI samples relative to healthy, and MI samples relative to CAD.") %>%
  kable_styling(full_width = T) %>%
  add_header_above(c(" " = 1, "CAD/Healthy" = 2, "MI/Healthy" = 2, "MI/CAD" = 2))
```

```{r Expall, echo=F, out.width="90%", fig.align='center', fig.cap="Expression profile of all miRNAs in two approaches in different sample classes."}
knitr::include_graphics("/Users/mehrdad/Projects/TBZ.Med/PhD. Project/Bioinformatics and ML/Article Materials/ML/Pics/Expressionforall.pdf")
```
